Parameter Inference\\with Bifurcation Diagrams

### 2021-05-28 docs/appendix.tex
Updates from Overleaf

Suppose there exists a one dimensional space curve $\mathcal{C(\theta)}$ embedded in $z\in\Reals^{N+1}$ whos geometry changes depending on input parameters $\theta\in\Reals^M$. This curve could be open or closed and changes in $\theta$ could change the curve topology as well. Let the function $\gamma_{\theta}:\Reals\rightarrow\Reals^{N+1}$ be a parameterisation of the position vector along the curve within a fixed domain $s\in\mathcal{S}$. Note that the choice of parameterisation is arbitrary and our results should not depend on this choice. Furthermore, if we parametrise the curve $\mathcal{C}(\theta)$ with respect to a fixed domain $\mathcal{S}$ the dependence on $\theta$ is picked up by the parameterisation $\gamma_{\theta}(s)$. We can write a line integral of any scalar function $L_{\theta}:\Reals^{N+1}\rightarrow\Reals$ on the curve as

>>>>

Suppose there exists a one dimensional space curve $\mathcal{C(\theta)}$ embedded in $z\in\Reals^{N+1}$ whose geometry changes depending on input parameters $\theta\in\Reals^M$. This curve could be open or closed and changes in $\theta$ could change the curve topology as well. Let the function $\gamma_{\theta}:\Reals\rightarrow\Reals^{N+1}$ be a parameterisation of the position vector along the curve within a fixed domain $s\in\mathcal{S}$. Note that the choice of parameterisation is arbitrary and our results should not depend on this choice. Furthermore, if we parametrise the curve $\mathcal{C}(\theta)$ with respect to a fixed domain $\mathcal{S}$ the dependence on $\theta$ is picked up by the parameterisation $\gamma_{\theta}(s)$. We can write a line integral of any scalar function $L_{\theta}:\Reals^{N+1}\rightarrow\Reals$ on the curve as

### 2021-05-28 docs/article.tex
Updates from Overleaf

Inverse problems \cite{Abdulla2009InverseBiology} arise in biology and engineering in settings when the model is not fully known and the desire is to match model behaviour to a given set of observations. This helps systematically guide both model and experimental design. The collected observations, however, are often indirectly proportional to the variables defined in the model; in microscopy, for example, data are reported in arbitrary fluorescence units allowing the observer to shift and scale data arbitrarily. The model, on the other hand, describes the dynamics of concentrations of relevant proteins. Furthermore, the observed qualitative changes in response to changes in experimental conditions are more robust and reproducible across studies than the quantitative details. For example, several studies are likely to agree that the human immune system activates above a threshold concentration of a pathogen and deactivates at a lower threshold concentration, but may disagree on the exact quantities of the thresholds or the magnitudes of the immune response. Bifurcation theory provides us a framework for studying these qualitative changes in a manner that is independent of quantitative details \cite{Kuznetsov2004TopologicalSystems}. The emerging picture suggests that identification of the qualitative behaviour -- the bifurcation diagram -- should precede any attempt at inferring other properties of a system \cite{Stumpf2019ParameterBifurcations}.

Inferring the parameters of a model directly from a bifurcation diagram is difficult because it is not obvious how one should change the parameters to create a bifurcation. It could even be impossible for the model to bifurcate in the manner desired. Several approaches exist to place bifurcations to desired locations once a manifold is present \cite{Iwasaki1997AnType,Lu2006InverseSystems,Dobson2004DistanceBifurcations} yet typically resort to sampling techniques to search for them in the first place \cite{Chickarmane2005BifurcationTool,Conrad2006BifurcationClock}. Progress has been made is cases where model structure and stability conditions are used to refine the search space \cite{Otero-Muras2018Optimization-basedModels,Otero-Muras2014ACurves} yet the resulting objectives are still not explicit in the bifurcation targets and also not differentiable. A scalable method for navigating the space of bifurcation diagrams would enable design of differential equations with high-level qualitative constraints. Furthermore one could begin organising models according to qualitatively distinct behaviours.

Back-propagation through differential equation solvers has been a breakthrough over the past couple of years \cite{Chen2018NeuralEquations,Rackauckas2019DiffEqFlux.jl-AEquations} that enabled scalable parameter inference for differential equations from trajectory data. Although one could use trajectory data to create the aforementioned qualitative constraints \cite{Ranciati2017BayesianParameters,Khadivar2021LearningBifurcations} this would entail over-constraining information originating from the kinetics and dynamical transients of the model. Furthermore, such data usually does not contain sufficient information about dynamical transients in order to identify kinetic parameters. Techniques for back-propagating through implicit equation solvers have also been developed \cite{Look2020DifferentiableLayers,Bai2019DeepModels} although to the best of the authors' knowledge have not been applied to bifurcation diagrams at the time of writing this paper.

In this work, we propose a gradient-based semi-supervised approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. The bifurcation diagram encodes high-level qualitative information defined by state space structures, rather than kinetics. We apply the strategy of implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients. To compute the diagram we use a predictor-corrector method called deflated pseudo-arclength continuation \cite{Farrell2016TheDiagrams,Veltz2019PseudoArcLengthContinuation.jl} developed for partial differential equations. In the case of partial differential equations, the computational complexity of calculating even a single bifurcation diagram can be unbounded since superpositions of localised solutions may give rise to uncountably many branches \cite{Avitabile2010ToEquation}. The complexity of computing a single branch, however, is bounded by the complexities of the chosen eigenvalue solver and corrector, and further decreased by adaptive stepping procedures \cite{Aruliah2016AlgorithmContinuation}.

@@

For a given set of parameters $\theta$ one could compute the set of predicted bifurcations $\predictions(\theta)$ using parameter continuation methods \cite{Veltz2019PseudoArcLengthContinuation.jl,Farrell2016TheDiagrams}. Our goal is to find optimal parameters $\theta^*$ that match predictions $\predictions(\theta^*)$ to specified targets $\targets$. We must design a suitable cost function $\loss$ so that
\begin{align}
    \theta^* := \mathrm{argmin}_{\theta} \loss(\theta|\targets)
    \label{eq:optimal-theta}
\end{align}
The optimal $\theta^*$ is not expected to always be unique, but is in general a manifold representing the space of qualitatively equivalent models. Ideally, the cost function $\loss$ should reward $\theta$ for which the number of predicted bifurcations is equal to the number of targets, $|\predictions(\theta)|=|\targets|$. This is especially important in the case where there are no predictions $|\predictions(\theta)|=0$.

@@

In order for predicted bifurcations $p(\theta)\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. A naive approach might take an average over the norms for each prediction-target pair. However, this can result in multiple predictions matching the same target. Therefore, we choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target (in cases where $|\predictions|\geq|\targets|$). Such an error term is only zero when each target is matched by at least one prediction
\begin{align}
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:error}
\end{align}

\subsubsection{Unsupervised term: encouraging bifurcations}
\begin{wrapfigure}{r}{7cm}
    \centering
    \includegraphics[width=\linewidth]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(s)$ and determinant $\Det$ along the arclength $s$ of two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars}
    \label{fig:measure}
\end{wrapfigure}

We can see from Figures \ref{fig:minimal-models} and definitions \eqref{eq:predictions} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$. Using these quantities we can define a positive semi-definite measure $\measure(s)$ of zero crossings in the determinant along the bifurcation curve, which we define as 
\begin{equation}
    \measure(s):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{equation}
This measure is maximal at bifurcations and has finite gradients in non-bifurcating regimes (Figure \ref{fig:measure}). More specifically, the measure $\measure(s)$ is one at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in the directional derivative $\frac{d}{ds}\Det$ going to zero and hence the measure $\measure(s)$ going to zero for each turning point.

To quantify the overall propensity for bifurcations over a diagram, we define the total bifurcation measure $\Psi(\theta)$ as
\begin{equation}
    \Psi(\theta):=\frac{
        \int_{\rates(u,p)=0}\!
        \measure(s)\,\mathrm{d}s
    }{
        \int_{\rates(u,p)=0}\!\
        \mathrm{d}s
    }
\end{equation}
Here we denote $\int_{\rates(u,p)=0}\mathrm{d}s$ as the sum of the line integrals in $(u,p)\in\Reals^{N+1}$ defined by the level set $\rates(u,p)=0$ with $s$ being an arbitrary parametrisation of the curves.

@@

As the determinant $\Det$ diverges we approach regimes far away from any closest bifurcation and hence $\measure(s)\rightarrow0$. We would still like to have non-zero gradients with respect to $\theta$ in this regime, and therefore $\measure(s)$ was designed to go to zero slowly. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the controlled condition region $p$ is densely packed with bifurcations. The measure $\measure(s)$ is also invariant under certain classes of transformations on $\rates(u,p)$ which do not change bifurcation locations, proofs of which are provided in Appendix \ref{appendix:measure}. The total measure $\Psi(\theta)$ is added to the supervised term as if it were a likelihood; this defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\,\,\lambda \log\Psi(\theta)+
    \frac{1}{|\targets|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:loss}
\end{align}

@@

The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached. Note that while individual bifurcations $p(\theta)$ depend smoothly on $\theta$, the total number of predictions $|\predictions|$ does not have gradient contributions with respect to $\theta$. We can safely drop the dependency in the prediction counter and now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ and total measure $\Psi(\theta)$
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient-loss}
\end{align}
In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation diagram $\rates(u,p)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:leibniz-rule}.

@@

\clearpage
\subsection{Complexity}

\begin{wrapfigure}{r}{7cm}
\centering
\includegraphics[width=\linewidth]{scaling}
\caption{Complexity scaling of calculating the gradient of the cost function. Calculations were performed on an Intel Core i7-6700HQ CPU @ 2.60GHz x 8 without GPU acceleration}
\label{fig:scaling}
\end{wrapfigure}

@@

@@

\section{Acknowledgements}
We would like to acknowledge Kieran Cooney for the fruitful conversations that helped guide the derivations and computational approach.

>>>>

Inverse problems \cite{Abdulla2009InverseBiology} arise in biology and engineering in settings when the model is not fully known and the desire is to match model behaviour to a given set of observations. This helps systematically guide both model and experimental design. The collected observations, however, are often indirectly proportional to the variables defined in the model. In fluorescence microscopy, a commonly used technique in biological studies, data are reported in arbitrary fluorescence units, allowing the observer to shift and scale data arbitrarily. Models of cellular networks, on the other hand, often seek to describe the dynamics of concentrations of relevant proteins \cite{Tyson2001NetworkPhysiology,Grant2020InterpretationCircuit}. Furthermore, the observed qualitative changes in response to changes in experimental conditions are more robust and reproducible across studies than the quantitative details. For example, several studies are likely to agree that the human immune system activates above a threshold concentration of a pathogen and deactivates at a lower threshold concentration, but may disagree on the exact quantities of the thresholds or the magnitudes of the immune response. Bifurcation theory provides us a framework for studying these qualitative changes in a manner that is independent of quantitative details \cite{Kuznetsov2004TopologicalSystems,Tyson2001NetworkPhysiology}. The emerging picture suggests that identification of the qualitative behaviour -- the bifurcation diagram -- should precede any attempt at inferring other properties of a system \cite{Stumpf2019ParameterBifurcations}.

Inferring the parameters of a model directly from a bifurcation diagram is difficult because it is not obvious how multiple  parameters in concert control the existence and position of a bifurcation. It could even be impossible for the model to bifurcate in the manner desired. Several approaches exist to place bifurcations to desired locations once a manifold is present \cite{Iwasaki1997AnType,Lu2006InverseSystems,Dobson2004DistanceBifurcations} yet typically resort to sampling techniques to search for them in the first place \cite{Chickarmane2005BifurcationTool,Conrad2006BifurcationClock}. Progress has been made is cases where model structure and stability conditions are used to refine the search space \cite{Otero-Muras2018Optimization-basedModels,Otero-Muras2014ACurves} yet the resulting objectives are still not explicit in the bifurcation targets and also not differentiable. In the emerging field of Scientific machine learning \cite{Rackauckas2017Differentialequations.jlJulia,Rackauckas2018ASolutions,Rackauckas2020UniversalLearning}, parameters of structured mechanistic models are favoured over flexible models in larger parameter spaces. A scalable method for navigating the space of bifurcation diagrams would enable design of differential equations with high-level qualitative constraints. Furthermore one could begin organising models according to qualitatively distinct behaviours.

Back-propagation through differential equation solvers has been a breakthrough over the past couple of years \cite{Chen2018NeuralEquations,Rackauckas2019DiffEqFlux.jlEquations} that enabled scalable parameter inference for differential equations from trajectory data. Although one could use trajectory data to create the aforementioned qualitative constraints \cite{Ranciati2017BayesianParameters,Khadivar2021LearningBifurcations} this would entail over-constraining information originating from the kinetics and dynamical transients of the model. Furthermore, such data usually does not contain sufficient information about dynamical transients in order to identify kinetic parameters. Techniques for back-propagating through implicit equation solvers have also been developed \cite{Look2020DifferentiableLayers,Bai2019DeepModels} although to the best of the authors' knowledge have not been applied to bifurcation diagrams at the time of writing this paper.

In this work, we propose a gradient-based semi-supervised approach for inferring the parameters of differential equation models that give rise to bifurcations at specified values of a control parameter. The bifurcation diagram encodes high-level qualitative information defined by state space structures, rather than kinetics. We apply the strategy of implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients. To compute the diagram we use a predictor-corrector method called deflated continuation \cite{Farrell2016TheDiagrams,Veltz2020BifurcationKit.jl} developed for partial differential equations. In the case of partial differential equations, the computational complexity of calculating even a single bifurcation diagram can be unbounded since superpositions of localised solutions may give rise to uncountably many branches \cite{Avitabile2010ToEquation}. The complexity of computing a single branch, however, is bounded by the complexities of the chosen eigenvalue solver and corrector, and further decreased by adaptive stepping procedures \cite{Aruliah2016AlgorithmContinuation}.

@@

For a given set of parameters $\theta$ one could compute the set of predicted bifurcations $\predictions(\theta)$ using parameter continuation methods \cite{Veltz2020BifurcationKit.jl,Farrell2016TheDiagrams}. Our goal is to find optimal parameters $\theta^*$ that match predictions $\predictions(\theta^*)$ to specified targets $\targets$. We must design a suitable cost function $\loss$ so that
\begin{align}
    \theta^* := \mathrm{argmin}_{\theta} \loss(\theta|\targets)
    \label{eq:optimal-theta}
\end{align}
The optimal $\theta^*$ is not expected to always be unique, but is in general a manifold representing the space of qualitatively equivalent models. Ideally, the cost function $\loss$ should reward $\theta$ for which the number of predicted bifurcations is equal to the number of targets, $|\predictions(\theta)|=|\targets|$. This is especially important in the case where there are no predictions $|\predictions(\theta)|=0$.

@@

In order for predicted bifurcations $p(\theta)\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. A naive approach might take an average over the norms for each prediction-target pair. However, this can result in multiple predictions matching the same target. Therefore, we choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target (in cases where $|\predictions|\geq|\targets|$):
\begin{equation}
    \error(\theta, \targets) = \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    %\label{eq:error}
\end{equation}
The supervised term is only zero when each target is matched by at least one prediction.

\subsubsection{Unsupervised term: encouraging bifurcations}

\begin{wrapfigure}{r}{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(s)$ and determinant $\Det$ along the arclength $s$ of two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars.}
    \label{fig:measure}
\end{wrapfigure}

We can see from Figures \ref{fig:minimal-models} and definitions \eqref{eq:predictions} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$. Using these quantities we can define a positive semi-definite measure $\measure(s)$ of zero crossings in the determinant along the bifurcation curve, which we define as 
\begin{equation}
    \measure(s):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{equation}

The bifurcation measure $\measure(s)$ is maximal at bifurcations and has finite gradients in non-bifurcating regimes (Figure \ref{fig:measure}). More specifically, the measure $\measure(s)$ is one at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in the directional derivative $\frac{d}{ds}\Det$ going to zero and hence the measure $\measure(s)$ going to zero for each turning point (see Figure \ref{fig:measure}). As the determinant $\Det$ diverges, we approach regimes far away from any closest bifurcation and hence $\measure(s)\rightarrow0$. But we would still like to have non-zero gradients with respect to $\theta$ in this regime, and so the measure is designed to go to zero slowly. Finally, we note that $\measure(s)$ has the desirable property of being invariant to certain classes of transformations on $\rates(u,p)$ that do not change bifurcation locations (details provided in Appendix \ref{appendix:measure}).

@@

Equipped with a measure that quantifies the appearance of bifurcations along a bifurcation arc, we now define the total measure for a bifurcation diagram as
\begin{equation}
    \Psi(\theta):=\frac{
        \int_{\rates(u,p)=0}\!
        \measure(s)\,\mathrm{d}s
    }{
        \int_{\rates(u,p)=0}\!\
        \mathrm{d}s
    }.
\end{equation}
Here we denote $\int_{\rates(u,p)=0}\mathrm{d}s$ as the sum of the line integrals in $(u,p)\in\Reals^{N+1}$ defined by the level set $\rates(u,p)=0$ with $s$ being an arbitrary parametrisation of the curves. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the controlled condition region $p$ is densely packed with bifurcations.

The total measure $\Psi(\theta)$ is added to the supervised term as if it were a likelihood. This defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\,\,\lambda \log\Psi(\theta)+
    \error(\theta, \targets),
    \label{eq:loss}
\end{align}
where we introduce the hyperparameter $\lambda$ to enable control over the relative contribution of the unsupervised term. The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached.

@@

To make use of gradient-based optimisers to locate desired bifurcation diagrams, we here show how to differentiate the cost function.
First, we note that while individual bifurcations $p(\theta)$ depend smoothly on $\theta$, the total number of predictions $|\predictions|$ does not have gradient contributions with respect to $\theta$. Therefore, we can safely drop the dependency in the prediction counter and now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ (within the definition of $\error$) and the total measure $\Psi(\theta)$. Therefore, 
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient-loss}
\end{align}

In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation diagram $\rates(u,p)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:leibniz-rule}.

@@

%\clearpage
\subsection{Complexity}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
\centering
\includegraphics[width=0.48\textwidth]{scaling}
\caption{Complexity scaling of calculating the gradient of the cost function. Calculations were performed on an Intel Core i7-6700HQ CPU @ 2.60GHz x 8 without GPU acceleration}
\label{fig:scaling}
\end{wrapfigure}

@@

% \section{Acknowledgements}
% We would like to acknowledge Kieran Cooney for the fruitful conversations that helped guide the derivations and computational approach.
